Double White Dwarf Mergers on Adaptive Meshes\\ I. Methodology and Code Verification

### 2015-08-14 papers/wdmerger_I/paper.tex
finish comments on the main paper

Another that we consider is the energy conservation of the system. Recalling \autoref{sec:gravity_hydro_coupling}, 
there are several different methods of applying the gravitational source term to the hydrodynamics equations. In \castro\ we presently 
have four options, controlled by the parameter {\tt castro.grav\_source\_type}, which we will shortern to {\tt gs} for the present 
discussion. ${\tt gs} = 1$ is the standard cell-centered source term for gravity. ${\tt gs} = 2$ is a slight variation on this scheme 
that determines the value of the energy source term after the momentum source term has been applied; the former approach uses 
the uncorrected momenta in calculating $\rho \mathbf{u} g$, which we have found to be more accurate. ${\tt gs} = 3$ is entirely 
different: after calculating the new momenta, we reset the total energy to be equal to the internal energy plus the 
kinetic energy. This has the virtue of ensuring that there is no conflict due to discretization between the momentum and 
energy equations, and also correctly ensuring that the gravitational force does not directly change the internal energy---and thus the temperature---of the fluid. However, it explicitly sacrifices total energy conservation. ${\tt gs} = 4$ is 
the new conservative method of evaluating the energy source terms at cell faces. The results for the change in energy 
after a single orbit are seen in the first column of \autoref{table:sources}. The first two versions give reasonable and 
similar levels of energy conservation. The third has total energy changes on the order of 100\%, but this itself 
does not have a severe effect on the dynamics because in this scheme the total energy variable is effectively a placeholder 
value of the kinetic energy plus internal energy, rather than being evolved directly. The last scheme is nearly two orders 
of magnitude better in energy conservation, justifying the effort in varying the scheme.

@@

We are most interested in the stability of these systems over long timescales. To this end, we consider the same systems as above, 
but evolve them for 25 orbital periods. In \autoref{fig:circular_orbit_comparison} we illustrate the evolution of these systems 
by plotting the center of mass locations of the white dwarfs on the orbital ($xy$) plane. For the equal mass case in the inertial 
reference frame, the curves essentially fall on top of each other, indicating that the stars are indeed orbiting at the initial 
distance without substantial deviation. In the unequal mass case, the stars now orbit at different distances but do not much 
deviate from their initial distance. We do observe a secular decline in the center of mass of these stars, indicating that 
eventually the stellar orbits will degrade due to numerical error even without intervention. The co-rotating frame is different.
For clarity of visualization, we rotate these results back into the inertial frame before displaying their orbits.
The stellar orbit in the equal mass case does not show the same secular decline, and the system appears to be stable. However, 
the cost is that the oscillations about this equilibrium are slightly larger than in the inertial frame. The unequal mass case displays 
this effect starkly: while the average distance is not declining, there are huge oscillations about this average. In addition, 
the stars move substantially within the rotating frame itself, yielding the chaotic-looking orbits in the figure. The oscillations 
grow with time and will eventually cause the stars to merge; however, the stars would eventually merge even in the inertial frame 
case, and at a faster rate than the equal mass case. This suggests that this mass ratio is too large to ignore tidal effects, and 
that the tidal effects play out much differently in the rotating frame.

@@

Turning to the conservation properties of the system, we examine as fairly typical cases the equal mass system 
in the inertial frame for energy conservation (\autoref{fig:energy_conservation_equal}), and the unequal mass 
system in the rotating frame for angular momentum conservation (\autoref{fig:angular_momentum_conservation_unequal}). 
Angular momentum is conserved to within a few percent over the 25 orbits, while energy conservation is about 
an order of magnitude better. We note that while this is already a fairly good level of energy conservation, it is not 
nearly as good as the results of \citet{marcello:2012}. This is because we reset the internal energy to a level 
corresponding to our temperature floor when it goes negative, while \citeauthor{marcello:2012} do not reset and 
instead ignore the internal energy if it is negative. The resets impose an artificial floor on our ability to 
conserve energy, but they only happen in low-density regions and do not much affect the large-scale dynamics. 
Meanwhile, angular momentum conservation is not quite as good. This is linked to the decline (or increase) in 
the size of the orbit. This implies that we ought to be careful in concluding that at these moderate resolutions 
we can safely evolve systems for many dozens of orbits; this needs to be verified to ensure that an observed 
inspiral and merger is physically (not numerically) motivated.

@@

We use OpenMP to accelerate the work associated with the boxes owned by each 
MPI task. Originally \castro\ used OpenMP to accelerate individual loops in the 
hydrodynamics routines, such as the piecewise-parabolic edge state reconstruction 
and the conservative flux update. However, there is a significant amount of 
overhead associated with generating a new OpenMP region at each of the many 
different loops in a hydrodynamics algorithm. This makes such a strategy 
sub-optimal for use on many-core processors and GPUs. We have recently switched 
to a tiling approach, where an OpenMP region is generated at the start of 
the hydrodynamics routine and the individual threads separately work on 
different partitions of each box. This results in much less overhead for 
the threading. In general we obtain more efficient simulations than 
could be obtained using MPI only, because there are fewer boxes and thus 
less communication for a given number of processor cores. We are currently 
developing an approach to evaluating the equation of state and nuclear reactions 
(and possibly the entire hydrodynamics algorithm) on GPUs, which will allow us
to take advantage of the significant computational resources embedded in GPUs on 
certain systems.

@@

To examine the parallel performance of \castro, we performed both strong scaling 
and weak scaling tests on the Blue Waters machine at the National Center for 
Supercomputing Applications. For the weak scaling test, whose results are 
shown in \autoref{fig:weak_scaling}, we ran a uniform grid binary white 
dwarf simulation for resolutions of $128^3$ zones through $2048^3$ zones. The 
number of processors was scaled with the number of zones so that each processor 
had the same amount of work; the smallest test used 8 processors and the largest 
used 32,768 (note that the number of processor cores on a Blue Waters node is twice 
the number of floating point units on that node). The test was run for 10 timesteps,
with each timestep including two Poisson solves and a hydrodynamics update (though 
for a uniform grid calculation we generally do not need to perform any multigrid 
iterations for the first Poisson solve in a timestep, since the density distribution 
has not changed since the end of the last timestep). We disabled plotfile and 
checkpoint writing, as well as calculation of diagnostic information (the latter 
can contribute to a significant fraction of the run time at large processor 
counts if computed every timestep). We computed the median wall time required per 
time step for each simulation, and then normalized this to the median time per 
timestep for the smallest simulation. We find excellent weak scaling through 
4,096 processors. At the largest run, the simulation time required is 
more than 1.5 times the amount required for the smallest simulation. 
This is due entirely to the increased cost of the multigrid Poisson solve 
in each timestep and this cannot be mitigated except by improving 
communication or computation efficiency in the multigrid solver. We observe 
that this weak scaling behavior with Poisson gravity is a significant 
improvement over the results presented in the first \castro\ paper.

@@

The strong scaling test we perform uses a grid setup similar to what we 
use for well-resolved binary simulations. With a single refined level, 
we have approximately $5 \times 10^7$ zones spread over $\sim 1000$ grids.
On a second refined level, there are $2 \times 10^9$ zones spread over 
$\sim 16000$ grids. We run a scaling test for both cases, with the 
highest processor count in each case chosen so that the number of 
MPI tasks is similar to the number of grids. There are no relative gains 
to be achieved from further parallelism. The results are found in 
\autoref{fig:strong_scaling}. We find excellent scaling for low numbers 
of processors, though the long wall time per timestep required for the 
smallest processor counts makes such a simulation infeasible in practice.
The best tradeoff between reasonable wall time and parallel efficiency 
seems to be when there are approximately 2--4 grids per processor. The 
scaling behavior worsens at the highest processor counts, but this is 
an expected consequence of processors becoming work-starved. In general 
we find very good strong scaling behavior in the regime we are presently 
interested in, simulating the early phases of a simulation at moderate 
resolution. The strong scaling behavior is acceptable, though not perfect, 
at very large processor counts when self-gravity is considered.

>>>>

Another that we consider is the energy conservation of the
system. Recalling \autoref{sec:gravity_hydro_coupling}, there are
several different methods of applying the gravitational source term to
the hydrodynamics equations. In \castro\ we presently have four
options, controlled by the parameter {\tt castro.grav\_source\_type},
which we will shortern to {\tt gs} for the present discussion. ${\tt
  gs} = 1$ is the standard cell-centered source term for
gravity. ${\tt gs} = 2$ is a slight variation on this scheme that
determines the value of the energy source term after the momentum
source term has been applied; the former approach uses the uncorrected
momenta in calculating $\rho \mathbf{u} g$, which we have found to be
more accurate. \MarginPar{which is more accurate?  Also, the ``standard approach'' is {\tt gs = 2} --- this is what is written, for exmaple, in the PPM paper}${\tt gs} = 3$ is entirely different: after calculating
the new momenta, we reset the total energy to be equal to the internal
energy plus the kinetic energy. This has the virtue of ensuring that
there is no conflict due to discretization between the momentum and
energy equations, and also correctly ensuring that the gravitational
force does not directly change the internal energy---and thus the
temperature---of the fluid. However, it explicitly sacrifices total
energy conservation. ${\tt gs} = 4$ is the new conservative method of
evaluating the energy source terms at cell faces. The results for the
change in energy after a single orbit are seen in the first column of
\autoref{table:sources}. The first two versions give reasonable and
similar levels of energy conservation. The third has total energy
changes on the order of 100\%, but this itself does not have a severe
effect on the dynamics because in this scheme the total energy
variable is effectively a placeholder value of the kinetic energy plus
internal energy, rather than being evolved directly. The last scheme
is nearly two orders of magnitude better in energy conservation,
justifying the effort in varying the scheme.

@@

We are most interested in the stability of these systems over long
timescales. To this end, we consider the same systems as above, but
evolve them for 25 orbital periods. In
\autoref{fig:circular_orbit_comparison} we illustrate the evolution of
these systems by plotting the center of mass locations of the white
dwarfs on the orbital ($xy$) plane. For the equal mass case in the
inertial reference frame, the curves essentially fall on top of each
other, indicating that the stars are indeed orbiting at the initial
distance without substantial deviation. In the unequal mass case, the
stars now orbit at different distances but do not much deviate from
their initial distance. We do observe a secular decline in the center
of mass of these stars, indicating that eventually the stellar orbits
will degrade due to numerical error even without intervention. The
co-rotating frame is different.  For clarity of visualization, we
rotate these results back into the inertial frame before displaying
their orbits.  The stellar orbit in the equal mass case does not show
the same secular decline, and the system appears to be
stable. However, the cost is that the oscillations about this
equilibrium are slightly larger than in the inertial frame. The
unequal mass case displays this effect starkly: while the average
distance is not declining, there are huge oscillations about this
average. In addition, the stars move substantially within the rotating
frame itself, yielding the chaotic-looking orbits in the figure. The
oscillations grow with time and will eventually cause the stars to
merge; however, the stars would eventually merge even in the inertial
frame case, and at a faster rate than the equal mass case. This
suggests that this mass ratio is too large to ignore tidal effects,
and that the tidal effects play out much differently in the rotating
frame. \MarginPar{this is interesting---has anyone else seen this?}

@@

Turning to the conservation properties of the system, we examine as
fairly typical cases the equal mass system in the inertial frame for
energy conservation (\autoref{fig:energy_conservation_equal}), and the
unequal mass system in the rotating frame for angular momentum
conservation (\autoref{fig:angular_momentum_conservation_unequal}).
Angular momentum is conserved to within a few percent over the 25
orbits, while energy conservation is about an order of magnitude
better. We note that while this is already a fairly good level of
energy conservation, it is not nearly as good as the results of
\citet{marcello:2012}. This is because we reset the internal energy to
a level corresponding to our temperature floor when it goes negative,
while \citeauthor{marcello:2012} do not reset and instead ignore the
internal energy if it is negative. The resets impose an artificial
floor on our ability to conserve energy, but they only happen in
low-density regions and do not much affect the large-scale dynamics.
Meanwhile, angular momentum conservation is not quite as good.\MarginPar{``as good'' as what?}  This is
linked to the decline (or increase) in the size of the orbit. This
implies that we ought to be careful in concluding that at these
moderate resolutions we can safely evolve systems for many dozens of
orbits; this needs to be verified to ensure that an observed inspiral
and merger is physically (not numerically) motivated.

@@

We use OpenMP to accelerate the work associated with the boxes owned
by each MPI task. Originally \castro\ used OpenMP to accelerate
individual loops in the hydrodynamics routines, such as the
piecewise-parabolic edge state reconstruction and the conservative
flux update. However, there is a significant amount of overhead
associated with generating a new OpenMP region at each of the many
different loops in a hydrodynamics algorithm. This makes such a
strategy sub-optimal for use on many-core processors and GPUs. We have
recently switched to a tiling approach \MarginPar{ref Weiqun's paper}, where an OpenMP region is
generated at the start of the hydrodynamics routine and the individual
threads separately work on different partitions of each box. This
results in much less overhead for the threading. In general we obtain
more efficient simulations than could be obtained using MPI only,
because there are fewer boxes and thus less communication for a given
number of processor cores. We are currently developing an approach to
evaluating the equation of state and nuclear reactions (and possibly
the entire hydrodynamics algorithm) on GPUs, which will allow us to
take advantage of the significant computational resources embedded in
GPUs on certain systems.

@@

To examine the parallel performance of \castro, we performed both
strong scaling and weak scaling tests on the Blue Waters machine at
the National Center for Supercomputing Applications. For the weak
scaling test, whose results are shown in \autoref{fig:weak_scaling},
we ran a uniform grid binary white dwarf simulation for resolutions of
$128^3$ zones through $2048^3$ zones. \MarginPar{is this really weak scaling?} The number of processors was
scaled with the number of zones so that each processor had the same
amount of work; the smallest test used 8 processors and the largest
used 32,768 (note that the number of processor cores on a Blue Waters
node is twice the number of floating point units on that node). The
test was run for 10 timesteps, with each timestep including two
Poisson solves and a hydrodynamics update (though for a uniform grid
calculation we generally do not need to perform any multigrid
iterations for the first Poisson solve in a timestep, since the
density distribution has not changed since the end of the last
timestep). We disabled plotfile and checkpoint writing, as well as
calculation of diagnostic information (the latter can contribute to a
significant fraction of the run time at large processor counts if
computed every timestep). We computed the median wall time required
per time step for each simulation, and then normalized this to the
median time per timestep for the smallest simulation. We find
excellent weak scaling through 4,096 processors. At the largest run,
the simulation time required is more than 1.5 times the amount
required for the smallest simulation.  This is due entirely to the
increased cost of the multigrid Poisson solve in each timestep and
this cannot be mitigated except by improving communication or
computation efficiency in the multigrid solver. We observe that this
weak scaling behavior with Poisson gravity is a significant
improvement over the results presented in the first \castro\ paper.

@@

The strong scaling test we perform uses a grid setup similar to what we 
use for well-resolved binary simulations. With a single refined level, 
we have approximately $5 \times 10^7$ zones spread over $\sim 1000$ grids.
On a second refined level, there are $2 \times 10^9$ zones spread over 
$\sim 16000$ grids. We run a scaling test for both cases, with the 
highest processor count in each case chosen so that the number of 
MPI tasks is similar to the number of grids. There are no relative gains 
to be achieved from further parallelism. The results are found in 
\autoref{fig:strong_scaling}. We find excellent scaling for low numbers 
of processors, though the long wall time per timestep required for the 
smallest processor counts makes such a simulation infeasible in practice.
The best tradeoff between reasonable wall time and parallel efficiency 
seems to be when there are approximately 2--4 grids per processor. The 
scaling behavior worsens at the highest processor counts, but this is 
an expected consequence of processors becoming work-starved. In general 
we find very good strong scaling behavior in the regime we are presently 
interested in, simulating the early phases of a simulation at moderate 
resolution. The strong scaling behavior is acceptable, though not perfect, 
at very large processor counts when self-gravity is considered. \MarginPar{does this use the new MG initialization of $\Phi$ from the old step?}
